---
title: "MaxQuant contaminants"
author: "Charlotte Dawson"
output: html_notebook
---

```{r}
library(httr)
library(magrittr)
library(rvest)

## This script also calls functions from the following packages:
# Biostrings
# data.table
# curl

## Define some urls for GET requests
BASE <- "https://www.uniprot.org"
TOOL_ENDPOINT <- "/uploadlists/" # REST endpoint
KB_ENDPOINT <- "/uniprot/" # UniProt website search endpoint
UPARC_ENDPOINT <- "/uniparc/" # UniParc website search endpoint
```

## What are contaminants in MaxQuant?

Description.

## Investigating contaminants

### Subsubheading

```{r}
# download original MaxQuant contaminants FASTA if it does not already exist
if (!file.exists("mq_contaminants_original.fasta")) {
  curl::curl_download(
    url = "http://lotus1.gwdg.de/mpg/mmbc/maxquant_input.nsf/7994124a4298328fc125748d0048fee2/$FILE/contaminants.fasta", 
    destfile = "mq_contaminants_original.fasta"
  )
}

# load MaxQuant contaminants FASTA
con_fasta <- Biostrings::readAAStringSet("mq_contaminants_original.fasta")
```

How many sequences?

```{r}
con_fasta
```

 And what do the headers look like?
 
```{r}
head(names(con_fasta))
tail(names(con_fasta))
```
 
The headers show the protein sequences have come from multiple different 
databases. Also the formatting of the UniProt headers is not standard which
makes life difficult. We can extract the identifiers from the different
databases from each FASTA header.

```{r}
# extract all protein IDs from FASTA headers
ids <- regexec("^[^ ]+", names(con_fasta)) %>% 
  regmatches(names(con_fasta), .) %>% 
  unlist()
head(ids)
tail(ids)
```


In this notebook we will try to replicate the MaxQuant
contaminants FASTA with up-to-date headers. 

## UniProt sequences

```{r}
# extract UniProt accessions
up <- grep(
  "^[QPOA][A-Z,0-9]{5}", 
  ids, 
  value = TRUE
)
length(up)
```




## Non-UniProt sequences
