---
title: "MaxQuant contaminants"
author: "Charlotte Dawson"
output: html_notebook
---

```{r}
library(httr)
library(magrittr)
library(rvest)
library(Biostrings)

## This script also calls functions from the following packages:
# Biostrings
# data.table
# curl

## Define some urls for GET requests
BASE <- "https://www.uniprot.org"
TOOL_ENDPOINT <- "/uploadlists/" # REST endpoint
KB_ENDPOINT <- "/uniprot/" # UniProt website search endpoint
UPARC_ENDPOINT <- "/uniparc/" # UniParc website search endpoint

# base R str_extract
str_extract <- function(pattern, string) {
  regexec(pattern, string, perl = TRUE) %>% 
    regmatches(string, .) %>% 
    unlist()
}
```

## What are contaminants in MaxQuant?

Description.

## Investigating contaminants (245 sequences)

### Subsubheading

```{r}
# download original MaxQuant contaminants FASTA if it does not already exist
if (!file.exists("mq_contaminants_original.fasta")) {
  curl::curl_download(
    url = "http://lotus1.gwdg.de/mpg/mmbc/maxquant_input.nsf/7994124a4298328fc125748d0048fee2/$FILE/contaminants.fasta", 
    destfile = "mq_contaminants_original.fasta"
  )
}

# load MaxQuant contaminants FASTA
con_fasta <- Biostrings::readAAStringSet("mq_contaminants_original.fasta")
```

How many sequences?

```{r}
con_fasta
```

Any duplicated sequences?

```{r}
any(Biostrings::duplicated(con_fasta))
```

Remove duplicate sequence

```{r}
con_fasta <- con_fasta[!Biostrings::duplicated(con_fasta)]
con_fasta
```

 And what do the headers look like?
 
```{r}
head(names(con_fasta))
tail(names(con_fasta))
```
 
The headers show the protein sequences have come from multiple different 
databases. Also the formatting of the UniProt headers is not standard which
makes life difficult. We first separate the sequences depending on whether 
the header contains a UniProt accession or not.

```{r}
# UniProt sequences
up <- con_fasta[grep("^[QPOA][A-Z,0-9]{5}", names(con_fasta))]

# Non-Uniprot sequences
non <- con_fasta[grep("^[QPOA][A-Z,0-9]{5}", names(con_fasta), invert = TRUE)]
```

In this notebook we will try to replicate the MaxQuant
contaminants FASTA with up-to-date headers. 

## UniProt sequences (209 sequences)

Extract accessions.

```{r}
up_accessions <- str_extract("^[QPOA][A-Z,0-9]{5}", names(up))
head(up_accessions)
```

Use UniProtKB accessions to query UniParc for sequences.

```{r}
payload <- list(
  query = paste(up_accessions, collapse = " OR "),
  format = "fasta"
)

tmp <- tempfile()
response <- GET(url = paste0(BASE, UPARC_ENDPOINT), query = payload, config = write_disk(tmp))

if (response$status_code == 200) {
  up_uparc <- Biostrings::readAAStringSet(tmp)
  message(paste("Input length:", length(up_accessions)))
  message(paste("Output FASTA length:", length(up_uparc)))
} else {
  stop("Something went wrong. Status code: ", response$status_code)
}
```

Of the `r length(up_uparc)` sequences we have obtained from UniParc, we
see which match the `up` sequences.

```{r}
up_good <- Biostrings::intersect(up_uparc, up)
up_good
```

We can see that `r length(up_good)` out of `r length(up)` sequences 
are accounted for. 

```{r}
up_bad <- setdiff(up, up_uparc)
up_bad
```

This leaves `r length(up_bad)` sequences which don't match the headers
they have been given in `contaminants.fasta`. 

Lets take these 'bad' accessions and search UniProtKB, instead of UniParc this
time.

```{r}
bad_accessions <- str_extract("^[QPOA][A-Z,0-9]{5}", names(up_bad))

payload <- list(
  query = paste(bad_accessions, collapse = " OR "),
  format = "fasta"
)

tmp <- tempfile()
response <- GET(url = paste0(BASE, KB_ENDPOINT), query = payload, config = write_disk(tmp))

if (response$status_code == 200) {
  bad_accessions_kb <- Biostrings::readAAStringSet(tmp)
  message(paste("Input length:", length(bad_accessions)))
  message(paste("Output FASTA length:", length(bad_accessions_kb)))
} else {
  stop("Something went wrong. Status code: ", response$status_code)
}
```

```{r}
up_good2 <- Biostrings::intersect(bad_accessions_kb, up_bad)
up_good2
```

Of the `r length(bad_accessions)` sequences from `up_can` that didn't match
their headers, we have now found found appropriate headers for 
`r length(up_good2)` of them.

```{r}
up_bad2 <- Biostrings::setdiff(up_bad, bad_accessions_kb)
up_bad2
```

The final `r length(up_bad2)` sequences I had to BLAST by hand in the
end as that was the quickest way to get metadata that actually matched the
sequences they were supposed to be referring to.

```{r}
blast <- read.delim("blast_table.tab")
str(blast)
```

```{r}

```


Blast table goes here...

Search for more informative headers.

Matching.

Fin.

## Non-UniProt sequences (35 sequences)

### RefSeq headers (6 sequences)

```{r}
# extract RefSeq accessions
refseq_accessions <- str_extract("(?<=REFSEQ:)[A-Z,0-9,_]+", names(non))
refseq_accessions
```

```{r}
payload <- list(
  query = paste(refseq_accession, collapse = " OR "),
  format = "fasta"
)

tmp <- tempfile()
response <- GET(url = paste0(BASE, UPARC_ENDPOINT), query = payload, config = write_disk(tmp))

if (response$status_code == 200) {
  refseq_uparc <- Biostrings::readAAStringSet(tmp)
  message(paste("Input length:", length(refseq_accessions)))
  message(paste("Output FASTA length:", length(refseq_uparc)))
} else {
  stop("Something went wrong. Status code: ", response$status_code)
}

```

```{r}
# get sequences with new UniParc headers
refseq <- Biostrings::intersect(refseq_uparc, non[grep("REFSEQ", names(non))])
refseq
```



### H-InvDb headers (3 sequences)

```{r}
# extract H-InvDB accessions
hinv_accessions <- str_extract("(?<=H-INV:)[A-Z,0-9]+", names(non))
hinv_accessions
```

```{r}
payload <- list(
  query = paste(hinv_accessions, collapse = " OR "),
  format = "fasta"
)

tmp <- tempfile()
response <- GET(url = paste0(BASE, UPARC_ENDPOINT), query = payload, config = write_disk(tmp))

if (response$status_code == 200) {
  hinv_uparc <- Biostrings::readAAStringSet(tmp)
  message(paste("Input length:", length(hinv_accessions)))
  message(paste("Output FASTA length:", length(hinv_uparc)))
} else {
  stop("Something went wrong. Status code: ", response$status_code)
}

```

```{r}
# get sequences with new UniParc headers
hinv <- Biostrings::intersect(hinv_uparc, non[grep("H-INV", names(non))])
hinv
```

### Ensembl headers (25 sequences)

```{r}
# extract Ensembl accessions
ens_accessions <- str_extract("(?<=ENSEMBL:)[A-Z,0-9,_]+", names(non))
ens_accessions
```

```{r}
payload <- list(
  query = paste(ens_accessions, collapse = " OR "),
  format = "fasta"
)

tmp <- tempfile()
response <- GET(url = paste0(BASE, UPARC_ENDPOINT), query = payload, config = write_disk(tmp))

if (response$status_code == 200) {
  ens_uparc <- Biostrings::readAAStringSet(tmp)
  message(paste("Input length:", length(ens_accessions)))
  message(paste("Output FASTA length:", length(ens_uparc)))
} else {
  stop("Something went wrong. Status code: ", response$status_code)
}

```

```{r}
# get sequences with new UniParc headers
ens_good <- Biostrings::intersect(ens_uparc, non[grep("ENSEMBL", names(non))])
ens_good
```

```{r}
ens_bad <- Biostrings::setdiff(non[grep("ENSEMBL", names(non))], ens_uparc)
ens_bad
```

Manual BLASTing.

```{r}
# obtain sequences with UniParc accession headers using old accessions 
payload <- list(
  query = paste(c(
    "UPI0000EBE2E6",
    "UPI0000EBCA6F",
    "UPI00005BDFA6",
    "UPI00005BC624"
  ), collapse = " OR "),
  format = "fasta"
)

tmp <- tempfile()
response <- GET(url = paste0(BASE, UPARC_ENDPOINT), query = payload, config = write_disk(tmp))

if (response$status_code == 200) {
  ens_bad_uparc <- Biostrings::readAAStringSet(tmp)
  message(paste("Input length:", length(ens_bad)))
  message(paste("Output FASTA length:", length(ens_bad_uparc)))
} else {
  stop("Something went wrong. Status code: ", response$status_code)
}
```

```{r}
ens <- Biostrings::union(ens_good, ens_bad_uparc)
ens
```

### Other headers (1 sequence)

```{r}
## Lastly, we will deal with any other proteins in the original FASTA
# extract anything else non-UniProt
oth_accession <- grep(
  "(ENSEMBL|REFSEQ|H-INV)", 
  names(non), 
  invert = TRUE, value = TRUE, perl = TRUE
)
oth_accession
```

```{r}
# it is streptavidin which we will just retrieve manually
payload <- list(
  query = "UPI000002B867",
  format = "fasta"
)

tmp <- tempfile()
response <- GET(url = paste0(BASE, UPARC_ENDPOINT), query = payload, config = write_disk(tmp))

if (response$status_code == 200) {
  oth <- Biostrings::readAAStringSet(tmp)
  message(paste("Input length:", length(oth_accession)))
  message(paste("Output FASTA length:", length(oth)))
} else {
  stop("Something went wrong. Status code: ", response$status_code)
}

oth
```

## Putting it all together

```{r}
con_fasta
```

```{r}
con_fasta_new <- c(up_good, up_good2, refseq, hinv, ens, oth)
length(con_fasta_new)
Biostrings::intersect(con_fasta_new, con_fasta)
```


